---
title: "Sweat gland outliers"
author: "Mari Myllymäki, Mikko Kuronen, Aila Särkkä"
date: "2023-09-19"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

In this example we analyse the sweat gland pattern of subject 203, a control.
The original data (video) have been collected by Dr. Kennedy’s group at the University of Minnesota by using the dynamic sweat test they have presented (Provitera et al. 2009. https://doi.org/10.1212/WNL.0b013e3181c7da4b ).
The point pattern analysed here was extracted by Kuronen et al. (2021, https://doi.org/10.1002/sim.8891) and
provided at the repository \url{https://github.com/mikkoku/SweatPaper} accompanying the article.

We obtained the sweat gland pattern of subject 203 as follows:

```{r, message=FALSE}
library(spatstat)
library(tidyverse)
```

```{r}
path <- "https://github.com/mikkoku/SweatPaper/raw/master/data"
glands <- read.table(file.path(path, "glands.csv"), header=TRUE, sep=",")
meta <- read.table(file.path(path, "meta.csv"), header=TRUE, sep=",")

s <- 17.5/2592 # scaling from pixels to millimeters
glands <- glands |> mutate(x=s*x, y=s*y)
meta <- meta |> mutate(across(3:6, ~ .x*s))

s203 <- glands |> filter(subjectid==203)
m203 <- meta |> filter(subjectid==203)

W <- owin(c(m203$x0, m203$x1),
          c(m203$y0, m203$y1))
X <- ppp(x=s203$x, y=s203$y, window=W)
par(mar=c(0,0,0,0), mgp=c(0,0,0))
plot(X, main="")
```

The nearest neighbor distance histogram shows some extraordinarily close pairs of points.

```{r}
par(mar=c(2.5,2.5,0.5,0.5))
hist(nndist(X), main="", breaks=seq(0, 2.1, by=0.05))
```

We plotted the point pattern and the points which had their nearest neighbor within distance 0.25 mm.

```{r}
nnlimit <- 0.25
par(mfrow=c(1,1), mar=c(0,0,0,0), mgp=c(0,0,0))
plot(X, main="")
points(subset(X, nndist(X) <= nnlimit), col=2, pch=16)
points(subset(X, nndist(X) <= nnlimit), col=1, pch=1)
```

To see what the effect of the close pairs to the model fit is, we removed 
one of each pair of points that had their nearest neighbor within distance 0.25 mm.

```{r}
Xr <- X
while(any(nndist(Xr) <= nnlimit)) {
  id <- which.min(nndist(Xr))
  Xr <- Xr[-id]
}
```

The pair correlation functions of the original and the filtered point patterns show difference in the short distances.

```{r}
# pcf of the original pattern
X.pcf <- pcf(X, correction="translate", r=seq(0, 3, length=101))
# pcf of the filtered pattern
Xr.pcf <- pcf(Xr, correction="translate", r=seq(0, 3, length=101))

par(mar=c(2.5,2.5,0.5,0.5))
plot(X.pcf, legend=FALSE, main="", ylim=c(0,1.4))
plot(Xr.pcf, add=TRUE, col=2)
```

We then investigated how the outlier pairs affect the parameters 
a softcore Gibbs point process model which has the density function
$$
f(\mathbf{x}) \propto \beta ^{n(\mathbf{x})}\exp\left(-\sum_{i<j}\left(\frac{\sigma}{\|x_i - x_j\|}\right)^{2/\kappa}\right),
$$
This model can be fitted by the R library `spatstat` (Baddeley et al. (2015)) using the `ppm` function.
It is possible to obtain confidence intervals for the model parameters using the `summary` function,
but only for internal parametrization.
We obtained confidence intervals for $\beta$ and $\sigma$ through transformation.

```{r}
parameter_estimates <- function(M) {
  # Transform internal parametrization to user friendly parametrization
  c1 <- coef(summary(M, fine=TRUE))
  if(c1[2,3] < 0) c1[2,3] <- 0
  rbind(beta=exp(c1[1,c(1,3,4)]), 
    sigma=M$interaction$par$sigma0*c1[2,c(1,3,4)]^(kappa/2))
}
kappa <- 0.2
M1 <- ppm(X~1, Softcore(kappa=kappa), correction="isotropic")
Mr <- ppm(Xr~1, Softcore(kappa=kappa), correction="isotropic")

M1
parameter_estimates(M1)

Mr
parameter_estimates(Mr)
```

Especially the interaction range parameter `sigma` shows a large difference between the 
point patterns.

## Pair correlation functions for all Control subjects in the data

Here the pair correlation functions of all five Control subjects in data show 
that the analysed pattern exhibits different behavior.
```{r}
dfpcf <- meta |> filter(group=="Controls") |>
  left_join(glands, by="subjectid") |>
  group_by(subjectid) |>
  reframe(pcf(ppp(x=x, y=y, window=owin(c(x0[1], x1[1]), c(y0[1], y1[1]))), 
              r=seq(0, 3, length=101)))
dfpcf |> filter(r != 0) |> ggplot() +
  geom_line(aes(x=r, y=trans, group=factor(subjectid), col=factor(subjectid))) +
  theme_minimal() +
  labs(y="pcf") +
  scale_color_discrete(type=grey.colors(2, end=0.7)[c(2,2,2,1,2)]) +
  guides(col="none")
```
